---
title: "Analysis of The Cancer Genome Atlas Lung Adenocarcinoma (LUAD) Samples Identifies Ecto-5’-nucleotidase (NT5E) as a Potential Driver of LUAD Tumorigenesis"
author: "2455225A"
date: "2023-08-17"
output: ioslides_presentation
#Created on R-Markdown and Transferred to Github
--- Installing Relevant Packages from Bioconductor- https://www.bioconductor.org/ and Preproccessing TCGA Data using the TCGAbiolinks Package

```{r setup, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager") +
  remove.packages("TCGAbiolinks")
BiocManager::install("TCGAbiolinks")
BiocManager::install("S4Arrays")
BiocManager::install("DelayedArray")
BiocManager::install("Biostrings")
library("TCGAbiolinks")
library("limma")
library("edgeR")
library("glmnet")
library("factoextra")
library("FactoMineR")
library("caret")
library("SummarizedExperiment")
library("gplots")
library("survival")
library("survminer")
library("RColorBrewer")
library("gProfileR")
library("genefilter")
library("TCGAretriever")
library("limma")

```
--- Selecting the RNA-seq data for Lung adenocarcinoma data in the TCGA 

```{r}
GDCprojects = getGDCprojects()
head(GDCprojects[c("project_id", "name")])
TCGAbiolinks:::getProjectSummary("TCGA-LUAD")
query_TCGA = GDCquery(
  project = "TCGA-LUAD",
  data.category = "Transcriptome Profiling", # parameter enforced by GDCquery
  experimental.strategy = "RNA-Seq",
  workflow.type = "STAR - Counts")
luad_res= getResults(query_TCGA)
colnames(luad_res)
head(luad_res$sample_type)
summary(factor(luad_res$sample_type))

query_TCGA = GDCquery(
  project = "TCGA-LUAD",
  data.category = "Transcriptome Profiling", # parameter enforced by GDCquery
  experimental.strategy = "RNA-Seq",
  workflow.type = "STAR - Counts",
  sample.type = c("Primary Tumor",'Solid Tissue Normal'),data.type = 'Gene Expression Quantification')


GDCdownload(query=query_TCGA)
```
--- Tables for TCGA-LUAD data
```{r}
table(tcga_data@colData$vital_status)
table(tcga_data@colData$tumor_stage)
table(tcga_data@colData$definition)
table(tcga_data@colData$tissue_or_organ_of_origin)
table(tcga_data@colData$gender)
table(tcga_data@colData$race)
```
---Saving tcga data
```{r}
saveRDS(object = tcga_data,
        file = "tcga_data.RDS",
        compress = FALSE)
tcga_data = readRDS(file = "tcga_data.RDS")
```

--- Using Limma code- adapdted from https://www.costalab.org/wp-content/uploads/2020/11/R_class_D3.html 
Changed to fit study
```{r}
limma_pipeline = function(
    tcga_data,
    condition_variable,
    reference_group=NULL){
  
  design_factor = colData(tcga_data)[, condition_variable, drop=T]
  
  group = factor(design_factor)
  if(!is.null(reference_group)){group = relevel(group, ref=reference_group)}
  
  design = model.matrix(~ group)
  
  dge = DGEList(counts=assay(tcga_data),
                samples=colData(tcga_data),
                genes=as.data.frame(rowData(tcga_data)))
  
  # filtering
  keep = filterByExpr(dge,design)
  dge = dge[keep,,keep.lib.sizes=FALSE]
  rm(keep)
  
  # Normalization (TMM followed by voom)
  dge = calcNormFactors(dge)
  v = voom(dge, design, plot=TRUE)
  
  # Fit model to data given design
  fit = lmFit(v, design)
  fit = eBayes(fit)
  
  # Show top genes
  topGenes = topTable(fit, coef=ncol(design), number=100, sort.by="p")
  
  return(
    list(
      voomObj=v, # normalized data
      fit=fit, # fitted model and statistics
      topGenes=topGenes # the 100 most differentially expressed genes
    )
  )
}
```
--- Using condition variables 'definition' for healthy or LUAD samples, also using AJCC Pathologic states
```{r}
limma_res = limma_pipeline(
  tcga_data=tcga_data,
  condition_variable="definition",
  reference_group="Solid Tissue Normal"
)
saveRDS(object = limma_res,
        file = "limma_.RDS",
        compress = FALSE)



limma_res_stage= limma_pipeline(
  tcga_data=tcga_data,
  condition_variable="ajcc_pathologic_stage",
  reference_group="Stage I"
)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.


--- Plotting PCA definition

```{r pressure}
 plot_PCA_definition = function(voomObj, condition_variable){
    group = factor(voomObj$targets[, condition_variable])
    pca = prcomp(t(voomObj$E))
    # Take PC1 and PC2 for the plot
    plot(pca$x[,1:2],col=group, pch=19,main = "PCA of Normal Tissue and Tumour Lung Samples", xlab = "PC1", ylab = "PC2")
 
    # include a legend for points
    legend("bottomleft", inset=.01, levels(group), pch=19, col=1:length(levels(group)))
    return(pca)
  }
  definition_res_pca = plot_PCA_definition(limma_res$voomObj, "definition")
```

--- Removing suffixes from AJCC states so there are Stage I,II etc rather than IA, IIB etc

```{r pressure}
 #removing prefixes of the stages ie stage IA, will become Stage I and so on
  ajcc_res_pca = function(voomObj, condition_variable) {
    group = factor(voomObj$targets[, condition_variable])
    
    # Combine "Stage IA" and "Stage IB" into "Stage I" in the factor variable
    group_stages <- group
    group_stages[group_stages %in% c("Stage IA", "Stage IB")] <- "Stage I"
    
    # Combine "Stage IIA" and "Stage IIB" into "Stage II" in the factor variable
    group_stages[group_stages %in% c("Stage IIA", "Stage IIB")] <- "Stage II"
    
    # Combine "Stage IIIA" and "Stage IIIB" into "Stage III" in the factor variable
    group_stages[group_stages %in% c("Stage IIIA", "Stage IIIB")] <- "Stage III"
    
    # Combine "Stage IV" into "Stage IV" in the factor variable
    group_stages[group_stages %in% c("Stage IV")] <- "Stage IV"
    
    # Update the levels of the factor variable
    group_stages <- factor(group_stages, levels = c("Stage I", "Stage II", "Stage III", "Stage IV"))
    
    pca = prcomp(t(voomObj$E))
    
    # Take PC1 and PC2 for the plot
    plot(pca$x[,1], pca$x[,2], col = group_stages, pch = 19, main = "PCA of AJCC Pathologic Stages", xlab = "PC1", ylab = "PC2")
    
    # Include a legend for points
    legend("bottomleft", inset = 0.01, levels(group_stages), pch = 19, col = 1:length(levels(group_stages)))  
    return(pca)
  }
  
  res_pcastages = plot_PCA(limma_res$voomObj, "ajcc_pathologic_stage")
  
  res_pca
  library(cowplot)
  combined_plot_PCA <- plot_grid(res_pca,res_pcastages)
  
  
  save_plot("combined_plotenrichr.tiff", combined_plotenrichr, base_width = 25, base_height = 15, dpi = 300)
 
```
--- Transpose and Save Limma_res
```{r}

  d_mat = as.matrix(t(limma_res$voomObj$E))

  d_resp = as.factor(limma_res$voomObj$targets$definition)
 saveRDS(object = limma_res,
        file = "limma_res.RDS",
        compress = FALSE) 
```
--- Correlation Plots/ Survival PLots and gene expression beeswarm
```{r}
saveRDS(object = limma_res,
        file = "limma_res.RDS",
        compress = FALSE)
tcga_data = readRDS(file = "tcga_data.RDS")
limma_res = readRDS(file = "limma_res.RDS")
```

--- Creating a Survival Variable
```{r}
clinical = tcga_data@colData
d_mat = t(limma_res$voomObj$E)
expr_df = limma_res$voomObj$genes
# Select only Primary Tumour

clin_df = clinical[clinical$definition == "Primary solid Tumor",
                   c("patient",
                     "vital_status",
                     "days_to_death",
                     "days_to_last_follow_up",
                     "gender",
                     "ajcc_pathologic_stage")]
clin_df$deceased = clin_df$vital_status == "Dead"
clin_df$overall_survival = ifelse(clin_df$deceased,
                                 clin_df$days_to_death,
                                 clin_df$days_to_last_follow_up)
Surv(clin_df$overall_survival, clin_df$deceased)
```
--- Creating Expression objects for individual genes of interest
```{r}
#NT5E-----------------------------------------------------  

dim(expr_df)
NT5E_id = expr_df['ENSG00000135318.12', "gene_id"]


NT5E_name = expr_df['ENSG00000135318.12', "gene_name"]

NT5Eexpr_diseased = d_mat[rownames(clin_df), NT5E_id]
NT5Eexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), NT5E_id]


#KRAS-----------------------------------------------------

KRAS_id =expr_df['ENSG00000133703.13', "gene_id"]
# also get the common gene name of the first row
KRAS_name =expr_df['ENSG00000133703.13', "gene_name"]

KRASexpr_diseased = d_mat[rownames(clin_df), KRAS_id]
KRASexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), KRAS_id]


# EGFR-----------------------------------------------------------------------------
EGFR_id =expr_df['ENSG00000146648.19', "gene_id"]
# also get the common gene name of the first row
EGFR_name =expr_df['ENSG00000146648.19', "gene_name"]

EGFRexpr_diseased = d_mat[rownames(clin_df), EGFR_id]
EGFRexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), EGFR_id]

#EREGE------------------------------------------------------------------------------------
EREG_id =expr_df['ENSG00000124882.4', "gene_id"]
# also get the common gene name of the first row
EREG_name =expr_df['ENSG00000124882.4', "gene_name"]

EREGexpr_diseased = d_mat[rownames(clin_df), EREG_id]
EREGexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), EREG_id]


#AREG---------------------------------------------------------------------------
AREG_id =expr_df['ENSG00000109321.11', "gene_id"]
# also get the common gene name of the first row
AREG_name =expr_df['ENSG00000109321.11', "gene_name"]

AREGexpr_diseased = d_mat[rownames(clin_df), AREG_id]
AREGexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), AREG_id]
#cd39-----------------------------------------------------------------------------
cd39_id =expr_df['ENSG00000138185.20', "gene_id"]
# also get the common gene name of the first row
cd39_name =expr_df['ENSG00000138185.20', "gene_name"]

cd39expr_diseased = d_mat[rownames(clin_df), cd39_id]
cd39expr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), cd39_id]

#A2a/A2b------------------------------------------------------------------------------------------
A2a_id =expr_df['ENSG00000128271.22', "gene_id"]
# also get the common gene name of the first row
A2a_name =expr_df['ENSG00000128271.22', "gene_name"]
A2aexpr_diseased = d_mat[rownames(clin_df), A2a_id]
A2aexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), A2a_id]

#A2b--------------------------------------------------------------------------------------
A2b_id =expr_df['ENSG00000170425.3', "gene_id"]
# also get the common gene name of the first row
A2b_name =expr_df['ENSG00000170425.3', "gene_name"]
A2bexpr_diseased = d_mat[rownames(clin_df), A2b_id]
A2bexpr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), A2b_id]





#HIPK1-----------------------------------------------------------------------------

HIPK1_id =expr_df['ENSG00000163349.22', "gene_id"]
# also get the common gene name of the first row
HIPK1_name =expr_df['ENSG00000163349.22', "gene_name"]

HIPK1expr_diseased = d_mat[rownames(clin_df), HIPK1_id]
HIPK1expr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), HIPK1_id]


#MAPK1 -------------------------------------------------------------------------------------------------------------

MAPK1_id = expr_df['ENSG00000100030.15', "gene_id"]
# also get the common gene name of the first row
MAPK1_name =expr_df['ENSG00000100030.15', "gene_name"]

MAPK1expr_diseased = d_mat[rownames(clin_df),MAPK1_id]
MAPK1expr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), MAPK1_id]

#ERBB GENES
ERBB2_id =expr_df['ENSG00000141736.14', "gene_id"]
# also get the common gene name of the first row
ERBB2_name =expr_df['ENSG00000141736.14', "gene_name"]

ERBB2expr_diseased = d_mat[rownames(clin_df),ERBB2_id]
ERBB2expr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), ERBB2_id]
#ERBB4
ERBB3_id =expr_df['ENSG00000065361.16', "gene_id"]

ERBB3_name =expr_df['ENSG00000065361.16', "gene_name"]

ERBB3expr_diseased = d_mat[rownames(clin_df),ERBB3_id]
ERBB3expr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), ERBB3_id]

#ERBB4

ERBB4_id =expr_df['ENSG00000178568.15', "gene_id"]

ERBB4_name =expr_df['ENSG00000178568.15', "gene_name"]

ERBB4expr_diseased = d_mat[rownames(clin_df),ERBB4_id]
ERBB4expr_healthy = d_mat[setdiff(rownames(d_mat), rownames(clin_df)), ERBB4_id]

```
--- Creating expression plots
```{r}
library(ggpubr)
library(tidyr)
library(plyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(cowplot)
library(ggplot2) 
```
--- Creating groups for facet wrap beesawrm
```{r}
group_order <- c("Diseased EGFR", "Healthy EGFR", "Diseased NT5E", "Healthy NT5E",
                 "KRAS Diseased", "KRAS Healthy", "HIPK1 Diseased", "HIPK1 Healthy",
                 "MAPK1 Diseased", "MAPK1 Healthy","Healthy EREG","Diseased EREG",
                 "Healthy AREG","Diseased AREG","Healthy cd39","Diseased cd39","Healthy A2A","Diseased A2A","Healthy A2B","Diseased A2B",
                 "Diseased ERBB2","Healthy ERBB2","Diseased ERBB3","Healthy ERBB3","Diseased ERBB4","Healthy ERBB4")


data_all <- data.frame(
  Expression = c(EGFRexpr_diseased, EGFRexpr_healthy, NT5Eexpr_diseased, NT5Eexpr_healthy,
                 KRASexpr_diseased, KRASexpr_healthy, HIPK1expr_diseased,
                 HIPK1expr_healthy, MAPK1expr_diseased, MAPK1expr_healthy,AREGexpr_diseased, AREGexpr_healthy, EREGexpr_diseased,
                 EREGexpr_healthy, cd39expr_diseased, cd39expr_healthy,ERBB2expr_diseased, ERBB2expr_healthy,
                 ERBB3expr_diseased, ERBB3expr_healthy,ERBB4expr_diseased, ERBB4expr_healthy),
  Group = rep(c("EGFR", "EGFR", "NT5E", "NT5E",
                "KRAS", "KRAS", "HIPK1", "HIPK1",
                "MAPK1", "MAPK1","AREG","AREG","EREG","EREG","cd39","cd39","ERBB2","ERBB2","ERBB3","ERBB3","ERBB4","ERBB4"), 
              times = sapply(list(EGFRexpr_diseased, EGFRexpr_healthy, NT5Eexpr_diseased,
                                  NT5Eexpr_healthy, KRASexpr_diseased, KRASexpr_healthy,
                                  HIPK1expr_diseased, HIPK1expr_healthy,
                                  MAPK1expr_diseased, MAPK1expr_healthy,AREGexpr_diseased, AREGexpr_healthy, EREGexpr_diseased,
                                  EREGexpr_healthy, cd39expr_diseased, cd39expr_healthy,ERBB2expr_diseased, ERBB2expr_healthy,
                                  ERBB3expr_diseased, ERBB3expr_healthy,ERBB4expr_diseased, ERBB4expr_healthy), length)),

  Group = rep(c("Diseased", "Healthy", "Diseased", "Healthy",
                "Diseased", "Healthy", "Diseased", "Healthy",
                "Diseased", "Healthy", "Diseased", "Healthy",
                "Diseased", "Healthy", "Diseased", "Healthy",
                "Diseased", "Healthy","Diseased", "Healthy", "Diseased", "Healthy"
                ), 
              times = sapply(list(EGFRexpr_diseased, EGFRexpr_healthy, NT5Eexpr_diseased,
                                  NT5Eexpr_healthy, KRASexpr_diseased, KRASexpr_healthy,
                                  HIPK1expr_diseased, HIPK1expr_healthy,
                                  MAPK1expr_diseased, MAPK1expr_healthy,AREGexpr_diseased, AREGexpr_healthy, EREGexpr_diseased,
                                  EREGexpr_healthy, cd39expr_diseased, cd39expr_healthy,ERBB2expr_diseased, ERBB2expr_healthy,
                                  ERBB3expr_diseased, ERBB3expr_healthy,ERBB4expr_diseased, ERBB4expr_healthy),length)))

#################### Add another column, Facet column of each gene, do free y axis, listapply(),
#ifelse!, data_all(col=), make all figures cowplot, which label, prevents normalisation of data on plot

library(cowplot)
data_all$Facet<-ifelse(data_all$Group=="Diseased EGFR"|data_all$Group=="Healthy EGFR","EGFR",
                       ifelse(data_all$Group=="Diseased NT5E"|data_all$Group=="Healthy NT5E","NT5E",
                              ifelse(data_all$Group=="Diseased HIPK1"|data_all$Group=="Healthy HIPK1","HIPK1",
                                     ifelse(data_all$Group=="Diseased KRAS"|data_all$Group=="Healthy KRAS","KRAS",
                                            ifelse(data_all$Group=="Diseased MAPK1"|data_all$Group=="Healthy MAPK1","MAPK1",
                                                   ifelse(data_all$Group=="Diseased EREG"|data_all$Group=="Healthy EREG","EREG",
                                                          ifelse(data_all$Group=="Diseased AREG"|data_all$Group=="Healthy AREG","AREG",
                                                                 ifelse(data_all$Group=="Diseased CD39"|data_all$Group=="Healthy cd39","cd39",
                                                                        ifelse(data_all$Group=="Diseased ERBB2"|data_all$Group=="Healthy ERBB2","ERBB2",
                                                                               ifelse(data_all$Group=="Diseased ERBB3"|data_all$Group=="Healthy ERBB3","ERBB3",
                                                                                      ifelse(data_all$Group=="Diseased ERBB4"|data_all$Group=="Healthy ERBB4","ERBB4",NA
                                                                                             
                              
                              )))))))))))

 
gene_data <- data.frame(
  Gene = rep(c("EGFR", "NT5E", "KRAS", "HIPK1", "MAPK1", "AREG", "EREG", "cd39", "ERBB2", "ERBB3", "ERBB4"), each = 2),
  Group = rep(c("Diseased", "Healthy"), times = 11),
  Expression = c(EGFRexpr_diseased, EGFRexpr_healthy, NT5Eexpr_diseased, NT5Eexpr_healthy,
                 KRASexpr_diseased, KRASexpr_healthy, HIPK1expr_diseased,
                 HIPK1expr_healthy, MAPK1expr_diseased, MAPK1expr_healthy,
                 AREGexpr_diseased, AREGexpr_healthy, EREGexpr_diseased,
                 EREGexpr_healthy, cd39expr_diseased, cd39expr_healthy,
                 ERBB2expr_diseased, ERBB2expr_healthy,
                 ERBB3expr_diseased, ERBB3expr_healthy, ERBB4expr_diseased, ERBB4expr_healthy)
)

unique(data_all$Group.2)

```
---Create comparisons to get P-values
```{r}
data_all$Group.2<-paste0(data_all$Group,data_all$Group.1)
unique(data_all$Group.2)

comparisons<-list(c("EGFRDiseased","EGFRHealthy"),
                  c("NT5EDiseased","NT5EHealthy"),
                  c("KRASDiseased","KRASHealthy"),
                  c("HIPK1Diseased","HIPK1Healthy"),
                  c("MAPK1Diseased","MAPK1Healthy"),
                  c("AREGDiseased","AREGHealthy"),
                  c("EREGDiseased","EREGHealthy"),
                  c("cd39Diseased","cd39Healthy"),
                  c("ERBB2Diseased","ERBB2Healthy"),
                  c("ERBB3Diseased","ERBB3Healthy"),
                  c("ERBB4Diseased","ERBB4Healthy")
                  )
group2_labels <- c( "EGFR"="EGFR",
                    "NT5E"="NT5E",
                    "KRAS"="KRAS",
                    "HIPK1"="HIPK1",
                    "MAPK1"="MAPK1",
                    "AREG"="AREG",
                    "EREG" ="EREG",
                    "cd39"="cd39",
                    "ERBB2"="ERBB2",
                    "ERBB3"="ERBB3",
                    "ERBB4"="ERBB4")
```

```{r}
library(ggbeeswarm)


data_all$ColorGroup <- factor(data_all$Group.1, levels = c("Diseased", "Healthy"))
data_all$ColorGroup
P1<-ggplot(data_all, aes(x = Group.2, y = Expression,fill=Group.1,col=Group.1)) +
  geom_quasirandom(dodge.width = 1) +
  scale_x_discrete(labels = NULL)+
  ylab("Z Score of Gene Expression") +
   # stat_compare_means(data = data_all, comparisons = comparisons, method = "wilcox", paired = FALSE)+ 
  ggtitle("Distribution of Gene Expression") +
  facet_wrap(~ Group, scales = "free")+
  theme(axis.text.x = element_blank()) +
  theme_cowplot() 
   # NB: stat_compare_means(data = data_all, comparisons = comparisons, method = "wilcox", paired = FALSE) could only be ran without Facet, so P.values were taken manualy and a table was constructed next to the figure containing all P.values of comparisons.

```
---Correaltion Plots- Diseased Using pre-defined gene healthy data
```{r}
library(ggcorrplot)
NT5ECOMB<-matrix(c(expr_diseased,KRASexpr_diseased,EGFRexpr_diseased,MAPK1expr_diseased,HIPK1expr_diseased
                   ,ERBB2expr_diseased,ERBB3expr_diseased,ERBB4expr_diseased,AREGexpr_diseased,EREGexpr_diseased,cd39expr_diseased,A2aexpr_diseased,A2bexpr_diseased),ncol=13)

colnames(NT5ECOMB)<-c( 'NT5E ','KRAS ',
                       'EGFR ','ERK ','HIPK1 ','ERBB2','ERBB3','ERBB4','AREG','EREG','CD39','A2a','A2b')

p.NT5EmatALLCOMBD=p.mat<-cor_pmat(NT5ECOMB)
corr_matNT5ECOMB=cor(NT5ECOMB,method='s')
corrplot(corr_matNT5ECOMB,method = 'square')
cordiseasedplot<-ggcorrplot(corr_matNT5ECOMB,hc.order=F,
           title = 'Diseased Gene Expression Spearman Correlation',
           p.mat = p.NT5EmatALLCOMBD,sig.level = 0.05, insig = c("blank"),type='upper',
           lab=T)+theme_cowplot()+ theme(axis.text.x = element_text(angle = 90))+labs(x = NULL)

```
--- Correlation plot diseased using pre-defined gene diseased data
```{r}
healthyNT5ECOMB<-matrix(c(expr_healthy,KRASexpr_healthy,EGFRexpr_healthy,MAPK1expr_healthy,HIPK1expr_healthy
                          ,ERBB2expr_healthy,ERBB3expr_healthy,ERBB4expr_healthy,AREGexpr_healthy,EREGexpr_healthy,cd39expr_healthy,A2aexpr_healthy,A2bexpr_healthy)
                        ,ncol=13)

colnames(healthyNT5ECOMB)<-c( 'NT5E ','KRAS ',
                              'EGFR ','ERK ','HIPK1 ','ERBB2','ERBB3','ERBB4','AREG','EREG','CD39','A2a','A2b'
)
View(healthyNT5ECOMB)
p.NT5EmatALLCOMBH=p.mat<-cor_pmat(healthyNT5ECOMB)
corr_matNT5ECOMB=cor(healthyNT5ECOMB,method='s')


corhealthyplot <- ggcorrplot(corr_matNT5ECOMB, hc.order = FALSE,
                             title = 'Healthy Gene Expression Spearman Correlation',
                             p.mat = p.NT5EmatALLCOMBH, sig.level = 0.05, insig = c("blank"), type = 'upper',
                             lab = TRUE) +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90))+labs(x = NULL) 
corhealthyplot
combined_plotcorr <- plot_grid(corhealthyplot,cordiseasedplot, nrow = 2,ncol = 2)


save_plot("Overall_Corr_Plots.tiff", combined_plotcorr, base_width = 25, base_height = 15, dpi = 300)
```


--- Survival Curves
```{r}
library("TCGAbiolinks")
library("limma")
library("edgeR")
library("glmnet")
library("factoextra")
library("FactoMineR")
library("caret")
library("SummarizedExperiment")
library("gplots")
library("survival")
library("survminer")
library("RColorBrewer")
library("gProfileR")
library("genefilter")
library('lattice')
```
--- Subsetting gene expressions based on quartiles, genes taking from code above using survival object
--- NT5E Survplot
```{r}
upper_quartile_NT5E <- quantile(clin_df$NT5E_value, probs = 0.75)
lower_quartile_NT5E <- quantile(clin_df$NT5E_value, probs = 0.25)
upper_quartile_NT5E
# Create a new column for gene expression quartiles
clin_df$quartileNT5E <- cut(clin_df$NT5E_value, breaks = c(min(clin_df$NT5E_value), lower_quartile_NT5E, upper_quartile_NT5E, max(clin_df$NT5E_value)),
                        labels = c("below lower quartile", "between quartiles", "above upper quartile"), include.lowest = TRUE)

subset_df <- clin_df[clin_df$quartileNT5E %in% c("above upper quartile", "below lower quartile"), ]

# Fit a survival model using the subsetted data
fit_subset <- survfit(Surv(overall_survival, deceased) ~ quartileNT5E, data = subset_df)
pval <- surv_pvalue(fit_subset, data = clin_df)$pval
# Plot the survival curve for upper and lower quartiles

NT5EONLYPPLOT<-ggsurvplot(fit_subset, data = clin_df, surv.median.line = "hv", pval = T, legend.title = "Gene Expression NT5E")
NT5EONLYPPLOT
```
--- EREG Survival Plot
```{r}
clin_df$EREG_value = d_mat[rownames(clin_df), EREG_id]

clin_df$EREG_value
# Calculate upper and lower quartiles
upper_quartile_EREG <- quantile(clin_df$EREG_value, probs = 0.75)
lower_quartile_EREG <- quantile(clin_df$EREG_value, probs = 0.25)
upper_quartile_EREG

# Create a new column for gene expression quartiles
clin_df$quartileEREG <- cut(clin_df$EREG_value, breaks = c(min(clin_df$EREG_value), lower_quartile_EREG, upper_quartile_EREG, max(clin_df$EREG_value)),
                            labels = c("below lower quartile", "between quartiles", "above upper quartile"), include.lowest = TRUE)

subset_df_EREG <- clin_df[clin_df$quartileEREG %in% c("above upper quartile", "below lower quartile"), ]

# Fit a survival model using the subsetted data
fit_subset_EREG <- survfit(Surv(overall_survival, deceased) ~ quartileEREG, data = subset_df_EREG)
pval <- surv_pvalue(fit_subset_EREG, data = clin_df)$pval
# Plot the survival curve for upper and lower quartiles

EREGONLYPPLOT<-ggsurvplot(fit_subset_EREG, data = clin_df, surv.median.line = "hv", pval = T, legend.title = "Gene Expression EREG")
EREGONLYPPLOT
```
---AREG Survival Plot
```{r}
 #Get the expression values for the selected gene
clin_df$AREG_value = d_mat[rownames(clin_df), AREG_id]

clin_df$AREG_value
# Calculate upper and lower quartiles
upper_quartile_AREG <- quantile(clin_df$AREG_value, probs = 0.75)
lower_quartile_AREG <- quantile(clin_df$AREG_value, probs = 0.25)
upper_quartile_AREG

# Create a new column for gene expression quartiles
clin_df$quartileAREG <- cut(clin_df$AREG_value, breaks = c(min(clin_df$AREG_value), lower_quartile_AREG, upper_quartile_AREG, max(clin_df$AREG_value)),
                            labels = c("below lower quartile", "between quartiles", "above upper quartile"), include.lowest = TRUE)
subset_df_AREG <- clin_df[clin_df$quartileAREG %in% c("above upper quartile", "below lower quartile"), ]

# Fit a survival model using the subsetted data
fit_subset_AREG <- survfit(Surv(overall_survival, deceased) ~ quartileAREG, data = subset_df_AREG)
pval <- surv_pvalue(fit_subset_EREG, data = clin_df)$pval
# Plot the survival curve for upper and lower quartiles

AREGONLYPPLOT<-ggsurvplot(fit_subset_AREG, data = clin_df, surv.median.line = "hv", pval = T, legend.title = "Gene Expression AREG")
AREGONLYPPLOT
                            
```
--- NT5E and AREG/EREG fit
```{r}
subset_df_NT5E_EREG_hi <- clin_df[clin_df$quartileNT5E == "above upper quartile" & clin_df$quartileEREG == "above upper quartile", ]
subset_df_NT5E_EREG_lo <- clin_df[clin_df$quartileNT5E == "below lower quartile" & clin_df$quartileEREG== "below lower quartile", ]
subset_df_NT5E_hi_EREG_lo <- clin_df[clin_df$quartileNT5E == "above upper quartile" & clin_df$quartileEREG == "below lower quartile", ]
subset_df_NT5E_lo_EREG_hi <- clin_df[clin_df$quartileNT5E == "below lower quartile" & clin_df$quartileEREG == "above upper quartile", ]
## AREG
subset_df_NT5E_AREG_hi <- clin_df[clin_df$quartileNT5E == "above upper quartile" & clin_df$quartileAREG == "above upper quartile", ]
subset_df_NT5E_AREG_lo <- clin_df[clin_df$quartileNT5E == "below lower quartile" & clin_df$quartileAREG== "below lower quartile", ]
subset_df_NT5E_hi_AREG_lo <- clin_df[clin_df$quartileNT5E == "above upper quartile" & clin_df$quartileAREG == "below lower quartile", ]
subset_df_NT5E_lo_AREG_hi <- clin_df[clin_df$quartileNT5E == "below lower quartile" & clin_df$quartileAREG == "above upper quartile", ]
# Fit survival models for the quartile combinations
fit_subset <- survfit(Surv(overall_survival, deceased) ~ quartile, data = subset_df)
subset_df_NT5E_AREG_hi
#EREG
fit_NT5E_EREG_hi <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_EREG_hi)
fit_NT5E_EREG_lo <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_EREG_lo)
fit_NT5E_hi_EREG_lo <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_hi_EREG_lo)
fit_NT5E_lo_EREG_hi <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_lo_EREG_hi)
#AREG
fit_NT5E_AREG_hi <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_AREG_hi)
fit_NT5E_AREG_lo <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_AREG_lo)
fit_NT5E_hi_AREG_lo <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_hi_AREG_lo)
fit_NT5E_lo_AREG_hi <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_NT5E_lo_AREG_hi)

### Works to here


# Getting p_values for combinations
combined_subset_df_EREG <- rbind(subset_df_NT5E_EREG_hi, subset_df_NT5E_EREG_lo)
combined_subset_df_EREG$Group <- factor(c(rep("NT5E_EREG_hi", nrow(subset_df_NT5E_EREG_hi)),
                                     rep("NT5E_EREG_lo", nrow(subset_df_NT5E_EREG_lo))))


# Plot the survival curves for the quartile combinations EREG

NT5E_EREG_hi_plot <- ggsurvplot(fit_NT5E_EREG_hi, data = subset_df_NT5E_EREG_hi, risk.table = TRUE,
                                 conf.int = TRUE, legend.title = "Gene Expression NT5Ehi EREGhi",
                                 legend.labs = c("NT5E_hi + EREG_hi"),
                                 surv.median.line = "hv")
NT5E_EREG_hi_plot
NT5E_EREG_lo_plot <- ggsurvplot(fit_NT5E_EREG_lo, data = subset_df_NT5E_EREG_lo, risk.table = TRUE,
                                 conf.int = TRUE, legend.title = "Gene Expression NT5Elo EREGlo",
                                 legend.labs = c("NT5E_lo + EREG_lo"),
                                 surv.median.line = "hv")
NT5E_EREG_lo_plot

NT5E_hi_EREG_lo_plot <- ggsurvplot(fit_NT5E_hi_EREG_lo, data = subset_df_NT5E_hi_EREG_lo, risk.table = TRUE, 
                                    conf.int = TRUE, legend.title = "Gene Expression NT5Ehi EREGlo",
                                    legend.labs = c("NT5E_hi + EREG_lo"),
                                    surv.median.line = "hv")
NT5E_hi_EREG_lo_plot

NT5E_lo_EREG_hi_plot <- ggsurvplot(fit_NT5E_lo_EREG_hi, data = subset_df_NT5E_lo_EREG_hi, risk.table = TRUE,
                                    conf.int = TRUE, legend.title = "Gene Expression NT5Elo EREGhi",
                                    legend.labs = c("NT5E_lo + EREG_hi")
)

NT5E_lo_EREG_hi_plot

# Plot the survival curves for the quartile combinations AREG
pvalNT5REGhi <- surv_pvalue(fit_NT5E_AREG_hi, data = subset_df_NT5E_AREG_hi)$pval

NT5E_AREG_hi_plot <- ggsurvplot(fit_NT5E_AREG_hi, data = subset_df_NT5E_AREG_hi, risk.table = TRUE,
                                 conf.int = TRUE, legend.title = "Gene Expression NT5Ehi AREGhi",
                                 legend.labs = c("NT5E_hi + AREG_hi"),
                                 surv.median.line = "hv")
NT5E_AREG_hi_plot
NT5E_AREG_lo_plot <- ggsurvplot(fit_NT5E_AREG_lo, data = subset_df_NT5E_AREG_lo, risk.table = TRUE,
                                 conf.int = TRUE, legend.title = "Gene Expression NT5Elo AREGlo",
                                 legend.labs = c("NT5E_lo + AREG_lo"),
                                 surv.median.line = "hv")
NT5E_AREG_lo_plot

NT5E_hi_AREG_lo_plot <- ggsurvplot(fit_NT5E_hi_AREG_lo, data = subset_df_NT5E_hi_AREG_lo, risk.table = TRUE, 
                                    conf.int = TRUE, legend.title = "Gene Expression NT5Ehi AREGlo",
                                    legend.labs = c("NT5E_hi + AREG_lo"),
                                    surv.median.line = "hv")
NT5E_hi_AREG_lo_plot

NT5E_lo_AREG_hi_plot <- ggsurvplot(fit_NT5E_lo_AREG_hi, data = subset_df_NT5E_lo_AREG_hi, risk.table = TRUE,
                                    conf.int = TRUE, legend.title = "Gene Expression NT5Elo AREGhi",
                                    legend.labs = c("NT5E_lo + AREG_hi")
)

NT5E_lo_AREG_hi_plot
clin_df$combined_quartile_EREGNT5E <- paste(clin_df$quartileNT5E, clin_df$quartileEREG, sep = "_")

combined_survfit_EREGNT5E <- survfit(Surv(overall_survival, deceased) ~ combined_quartile_EREGNT5E, data = clin_df)
combined_survfit_EREGNT5E$group <- rep(c("subset_df_NT5E_EREG_hi", "subset_df_NT5E_EREG_lo", "subset_df_NT5E_hi_EREG_lo ", "subset_df_NT5E_lo_EREG_hi"),
                                      each = nrow(clin_df))



subset_df_EREGNT5E <- clin_df[clin_df$quartileNT5E %in% c("above upper quartile", "below lower quartile") & clin_df$quartileEREG %in% c("above upper quartile", "below lower quartile"), ]

# Fit survival model for the selected quartile combinations
combined_survfit_EREGNT5E <- survfit(Surv(overall_survival, deceased) ~ quartileNT5E + quartileEREG, data = subset_df_EREGNT5E)
```
---AREG NT5E survivalplot
```{r}
combined_plotSurv_EREGNT5E <- ggsurvplot(combined_survfit_EREGNT5E, data = subset_df_EREGNT5E,
                                        #pval = p_val,
                                        legend.title = "Gene Expression Levels",
                                        legend.labs = c("NT5E_lo_EREG_lo", "NT5E_lo_EREG_hi", "NT5E_hi_EREG_lo", "NT5E_hi_EREG_hi"),
                                        surv.median.line = "hv",
                                        risk.table = TRUE)

combined_plotSurv_EREGNT5E


####AREG Survplot
clin_df$combined_quartile_AREGNT5E <- paste(clin_df$quartileNT5E, clin_df$quartileAREG, sep = "_")

combined_survfit_AREGNT5E <- survfit(Surv(overall_survival, deceased) ~ combined_quartile_AREGNT5E, data = clin_df)
combined_survfit_AREGNT5E$group <- rep(c("subset_df_NT5E_AREG_hi", "subset_df_NT5E_AREG_lo", "subset_df_NT5E_hi_AREG_lo ", "subset_df_NT5E_lo_AREG_hi"),
                                      each = nrow(clin_df))
clin_df$quartileAREG
subset_df_AREGNT5E <- clin_df[clin_df$quartileNT5E %in% c("above upper quartile", "below lower quartile") & clin_df$quartileAREG %in% c("above upper quartile", "below lower quartile"), ]
unique(subset_df_AREGNT5E$quartileAREG)
unique(subset_df_AREGNT5E$quartileNT5E)

# Fit survival model for the selected quartile combinations
combined_survfitE_AREGNT5E <- survfit(Surv(overall_survival, deceased) ~ quartileNT5E + quartileAREG, data = subset_df_AREGNT5E)
unique(combined_survfit_AREGNT5E$group)
unique(subset_df_AREGNT5E$com)

combined_plotSurv_AREGNT5E <- ggsurvplot(combined_survfitE_AREGNT5E, data = subset_df_AREGNT5E,
                                        #pval = p_val,
                                        legend.title = "Gene Expression Levels",
                                        legend.labs = c("NT5E_lo_AREG_lo", "NT5E_lo_AREG_hi", "NT5E_hi_AREG_lo", "NT5E_hi_AREG_hi"),
                                        surv.median.line = "hv",
                                        risk.table = TRUE)

combined_plotSurv_AREGNT5E
```
--- Coxph -hazard ratios for NT5E, EREG+NT5E survival curves
```{r}
library(lubridate)
library(ggsurvfit)
library(gtsummary)
library(tidycmprsk)
### NT5E EREG coxph
subset_df_EREGNT5E$Combined_Expression <- paste0("NT5E_", subset_df_EREGNT5E$NT5E_Expression, "_EREG_", subset_df_EREGNT5E$EREG_Expression)


# Perform Cox regression for NT5Elo & EREGhi vs. NT5Elo & EREGlo (Reference)
subset_df_EREGNT5E$Combined_Expression <- relevel(factor(subset_df_EREGNT5E$Combined_Expression), ref = "NT5E_lo_EREG_lo")

coxph(Surv(overall_survival, deceased) ~ Combined_Expression, data = subset_df_EREGNT5E,
                       subset = Combined_Expression %in% c("NT5E_lo_EREGlo","NT5E_lo_EREG_hi", "NT5E_lo_EREG_lo","NT5E_hi_EREG_lo","NT5E_hi_EREG_hi")) %>% 
  tbl_regression(exp = TRUE) 
#### NT5E coxph
coxph(Surv(overall_survival, deceased) ~ NT5E_Expression, data = subset_df_EREGNT5E) %>% 
  tbl_regression(exp = TRUE) 
##### P values between comparisons also obtained most significnact pValue plotted on graphs, repeated for AREG+NT5E, ERBB2+KRAS comparisons


```















---KRAS survivalplot
```{r}
clin_df$KRAS_value=expr_df[rownames(clin_df), KRAS_id]

clin_df$KRAS_value = d_mat[rownames(clin_df), KRAS_id]

View(clin_df$ERBB2_value)
# Calculate upper and lower quartiles
upper_quartile_KRAS <- quantile(clin_df$KRAS_value, probs = 0.75)
lower_quartile_KRAS <- quantile(clin_df$KRAS_value, probs = 0.25)

# Create a new column for gene expression quartiles
clin_df$quartile_KRAS <- cut(clin_df$KRAS_value, breaks = c(min(clin_df$KRAS_value), lower_quartile_KRAS, upper_quartile_KRAS, max(clin_df$KRAS_value)),
                        labels = c("below lower quartile", "between quartiles", "above upper quartile"), include.lowest = TRUE)

# Subset the data to include only the upper and lower quartiles
subset_df_quartiles_KRAS <- clin_df[clin_df$quartile_KRAS %in% c("above upper quartile", "below lower quartile"), ]

# Fit a survival model using the subsetted data
fit_quartiles_KRAS <- survfit(Surv(overall_survival, deceased) ~ quartile_KRAS, data = subset_df_quartiles_KRAS)
pvalKRAS <- surv_pvalue(fit_quartiles_KRAS, data = subset_df_quartiles_KRAS)$pval

KRASONLYPPLOT<-ggsurvplot(fit_quartiles_KRAS, data = subset_df_quartiles_KRAS, surv.median.line = "hv", pval = T, legend.title = "Gene Expression KRAS")
KRASONLYPPLOT
```
---ERBB2 Survivalplot
```{r}
clin_df$ERBB2_value=expr_df[rownames(clin_df), ERBB2_id]

clin_df$ERBB2_value = d_mat[rownames(clin_df), ERBB2_id]
View(clin_df$ERBB2_value)
# Calculate upper and lower quartiles
upper_quartile_ERBB2 <- quantile(clin_df$ERBB2_value, probs = 0.75)
lower_quartile_ERBB2 <- quantile(clin_df$ERBB2_value, probs = 0.25)

# Create a new column for gene expression quartiles
clin_df$quartile_ERBB2 <- cut(clin_df$ERBB2_value, breaks = c(min(clin_df$ERBB2_value), lower_quartile_ERBB2, upper_quartile_ERBB2, max(clin_df$ERBB2_value)),
                        labels = c("below lower quartile", "between quartiles", "above upper quartile"), include.lowest = TRUE)

# Subset the data to include only the upper and lower quartiles
subset_df_quartiles_ERBB2 <- clin_df[clin_df$quartile_ERBB2 %in% c("above upper quartile", "below lower quartile"), ]

# Fit a survival model using the subsetted data
fit_quartiles_ERBB2 <- survfit(Surv(overall_survival, deceased) ~ quartile_ERBB2, data = subset_df_quartiles_ERBB2)
pvalK <- surv_pvalue(fit_quartiles_ERBB2, data = subset_df_quartiles_ERBB2)$pval
subset_df_quartiles_ERBB2

ERBB2ONLYPPLOT<-ggsurvplot(fit_quartiles_ERBB2, data = subset_df_quartiles_ERBB2, surv.median.line = "hv", pval = T, legend.title = "Gene Expression ERBB2")
ERBB2ONLYPPLOT
```
---KRAS and ERBB2 Combined Survival Plots
```{r}
# Create quartile categories for KRAS and ERBB2
clin_df$NT5E_quartile <- cut(clin_df$KRAS_value, breaks = c(min(clin_df$KRAS_value), lower_quartile_KRAS, upper_quartile_KRAS, max(clin_df$KRAS_value)),
                             labels = c("KRAS_lo", "KRAS_between", "KRAS_hi"), include.lowest = TRUE)
clin_df$ERBB2_quartile <- cut(clin_df$ERBB2_value, breaks = c(min(clin_df$ERBB2_value), lower_quartile_ERBB2, upper_quartile_ERBB2, max(clin_df$ERBB2_value)),
                              labels = c("ERBB2_lo","ERBB2 between", "ERBB2_hi"), include.lowest = TRUE)

# Subset the data based on the quartile combinations
View(subset_df_KRAS_hi_ERBB2_lo)
clin_df$quartile_KRAS
subset_df_KRAS_ERBB2_hi <- clin_df[clin_df$quartile_KRAS == "above upper quartile" & clin_df$quartile_ERBB2 == "above upper quartile", ]
subset_df_KRAS_ERBB2_lo <- clin_df[clin_df$quartile_KRAS == "below lower quartile" & clin_df$quartile_ERBB2== "below lower quartile", ]
subset_df_KRAS_hi_ERBB2_lo <- clin_df[clin_df$quartile_KRAS == "above upper quartile" & clin_df$quartile_ERBB2 == "below lower quartile", ]
subset_df_KRAS_lo_ERBB2_hi <- clin_df[clin_df$quartile_KRAS == "below lower quartile" & clin_df$quartile_ERBB2 == "above upper quartile", ]

# Fit survival models for the quartile combinations
fit_subset <- survfit(Surv(overall_survival, deceased) ~ quartile, data = subset_df)
subset_df_KRAS_ERBB2_hi
fit_KRAS_ERBB2_hi <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_KRAS_ERBB2_hi)
fit_KRAS_ERBB2_lo <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_KRAS_ERBB2_lo)
fit_KRAS_hi_ERBB2_lo <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_KRAS_hi_ERBB2_lo)
fit_KRAS_lo_ERBB2_hi <- survfit(Surv(overall_survival, deceased) ~ 1, data = subset_df_KRAS_lo_ERBB2_hi)

# Getting p_values for combinations
combined_subset_df <- rbind(subset_df_KRAS_ERBB2_hi, subset_df_KRAS_ERBB2_lo)
combined_subset_df$Group <- factor(c(rep("KRAS_ERBB2_hi", nrow(subset_df_KRAS_ERBB2_hi)),
                                     rep("KRAS_ERBB2_lo", nrow(subset_df_KRAS_ERBB2_lo))))
p_value_KRAS_ERBB2 <- survdiff(Surv(overall_survival, deceased) ~ Group, data = combined_subset_df)$chisq[1]
p_value_KRAS_ERBB2

# Plot the survival curves for the quartile combinations
pvalKRASERBBhi <- surv_pvalue(fit_KRAS_ERBB2_hi, data = subset_df_KRAS_ERBB2_hi)$pval

KRAS_ERBB2_hi_plot <- ggsurvplot(fit_KRAS_ERBB2_hi, data = subset_df_KRAS_ERBB2_hi, risk.table = TRUE,
                                 conf.int = TRUE, legend.title = "Gene Expression KRAShi ERBB2hi",
                                 legend.labs = c("KRAS_hi + ERBB2_hi"),
                                 surv.median.line = "hv")
KRAS_ERBB2_hi_plot
KRAS_ERBB2_lo_plot <- ggsurvplot(fit_KRAS_ERBB2_lo, data = subset_df_KRAS_ERBB2_lo, risk.table = TRUE,
                                 conf.int = TRUE, legend.title = "Gene Expression KRASlo ERBB2lo",
                                 legend.labs = c("KRAS_lo + ERBB2_lo"),
                                 surv.median.line = "hv")
KRAS_ERBB2_lo_plot

KRAS_hi_ERBB2_lo_plot <- ggsurvplot(fit_KRAS_hi_ERBB2_lo, data = subset_df_KRAS_hi_ERBB2_lo, risk.table = TRUE, 
                                    conf.int = TRUE, legend.title = "Gene Expression KRAShi ERBB2lo",
                                    legend.labs = c("KRAS_hi + ERBB2_lo"),
                                    surv.median.line = "hv")
KRAS_hi_ERBB2_lo_plot

KRAS_lo_ERBB2_hi_plot <- ggsurvplot(fit_KRAS_lo_ERBB2_hi, data = subset_df_KRAS_lo_ERBB2_hi, risk.table = TRUE,
                                    conf.int = TRUE, legend.title = "Gene Expression KRASlo ERBB2hi",
                                    legend.labs = c("KRAS_lo + ERBB2_hi")
                                )

KRAS_lo_ERBB2_hi_plot
# Plot survival fits on the same graph
# Create a new column to identify quartile combinations
clin_df$combined_quartileERBBKRAS <- paste(clin_df$quartile_KRAS, clin_df$quartile_ERBB2, sep = "_")

combined_survfitERBBKRAS <- survfit(Surv(overall_survival, deceased) ~ combined_quartileERBBKRAS, data = clin_df)
combined_survfitERBBKRAS$group <- rep(c("subset_df_KRAS_ERBB2_hi", "subset_df_KRAS_ERBB2_lo", "subset_df_KRAS_hi_ERBB2_lo ", "subset_df_KRAS_lo_ERBB2_hi"),
                              each = nrow(clin_df))



subset_dfERBBKRAS <- clin_df[clin_df$quartile_KRAS %in% c("above upper quartile", "below lower quartile") & clin_df$quartile_ERBB2 %in% c("above upper quartile", "below lower quartile"), ]

# Fit survival model for the selected quartile combinations
combined_survfitERBBKRAS <- survfit(Surv(overall_survival, deceased) ~ quartile_KRAS + quartile_ERBB2, data = subset_dfERBBKRAS)

combined_plotSurvERBBKRAS <- ggsurvplot(combined_survfitERBBKRAS, data = subset_dfERBBKRAS,
                                #pval = p_val,
                                legend.title = "Gene Expression Levels",
                                legend.labs = c("KRAS_lo_ERBB2_lo", "KRAS_lo_ERBB2_hi", "KRAS_hi_ERBB2_lo", "KRAS_hi_ERBB2_hi"),
                                surv.median.line = "hv",
                                risk.table = TRUE)


combined_plotSurvERBBKRAS

```
---Differetial gene expression analysis bewteen NT5Ehi and NT5Elo clusters
```{r}
library(SingleCellExperiment)
library(enrichR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(matrixStats)
library(MatrixGenerics)
library(installr)
library("TCGAbiolinks")
library("limma")
library("edgeR")
library("glmnet")
library("factoextra")
library("FactoMineR")
library("caret")
library("SummarizedExperiment")
library("gplots")
library("survival")
library("survminer")
library("RColorBrewer")
library("gProfileR")
library("genefilter")
library(data.table)
```
--- Using NT5Ehi and NT5Elo data based on expression quartiles, upper and lower quartiles.
```{r}
# Calculate upper and lower quartiles
upperquartile <- quantile(tcga_gene_expression_matrix['ENSG00000135318.12', ], probs = 0.75, type = 7)
expr_samples_above_quartile_mat <- tcga_gene_expression_matrix[, tcga_gene_expression_matrix['ENSG00000135318.12', ] > upperquartile]



lowerquartile <- quantile(tcga_gene_expression_matrix['ENSG00000135318.12', ], probs = 0.25, type = 7)

expr_samples_below_quartile_mat <- tcga_gene_expression_matrix[, tcga_gene_expression_matrix['ENSG00000135318.12', ] < lowerquartile]
```
--- Using Limma Package for DEA, using NT5Ehi and NT5Elo as condition var. NT5Elo as reference
```{r}
#create a condition of hi and lo for the two matrices before combining as single assay in a SE
condition_var_hi <- rep("hi", ncol(expr_samples_above_quartile_mat))
condition_var_lo <- rep("lo", ncol(expr_samples_below_quartile_mat))
condition_var <- factor(c(condition_var_hi, condition_var_lo))
NT5E_se <- SummarizedExperiment(
  assays = SimpleList(combined = cbind(expr_samples_above_quartile_mat, expr_samples_below_quartile_mat)),
  colData = DataFrame(condition = condition_var)
)
NT5E_se$condition

# Creating a condition variable to allow a downstream DEA between hi and lo NT5E expression samples


design = model.matrix(~condition_var)
dge = DGEList(counts=cbind(expr_samples_above_quartile_mat, expr_samples_below_quartile_mat),
              samples=condition_var)

# filtering
keep = filterByExpr(dge,design)
dge = dge[keep,,keep.lib.sizes=FALSE]
rm(keep)

# Normalization (TMM followed by voom)
dge = calcNormFactors(dge)
v = voom(dge, design, plot=TRUE)

# Fit model to data given design
fit = lmFit(v, design)
fit = eBayes(fit)

summer.fit=decideTests(fit)
summary(summer.fit)

limma.resNT5E <- topTable(fit,coef=2,number=Inf)
limma.resNT5E$logFC
dim(limma.res)
saveRDS(object = limma.resNT5E,
         file = "limma.res.RDS",
         compress = FALSE)
limma.resNT5E = readRDS(file = "limma.res.RDS")
limma.resSolidTumour<-topTable(fit)
```
--- Getting Replacing ensembl IDs with gene symbols
```{r}
rownames(limma.res_high) <- sub("\\..*", "", rownames(limma.res_high))
rownames(limma.res_low) <- sub("\\..*", "", rownames(limma.res_low))

ens2symbol_high<- AnnotationDbi::select(org.Hs.eg.db,keys = rownames(limma.res_high),                                   columns="SYMBOL",keytype="ENSEMBL")

ens2symbol_low<- AnnotationDbi::select(org.Hs.eg.db,keys = rownames(limma.res_low),                                             columns="SYMBOL",keytype="ENSEMBL")

ens2symbol_high<- as_tibble(ens2symbol_high)

ens2symbol_low <- as_tibble(ens2symbol_low)


```
--- Getting ranked set of genes based on the moderated T statistic from the DEA for Gene set enrichment analysis (GSEA).
```{r}
###### Ranked genes T value
limma.res_high$t
res2_low <- limma.res_low %>% 
  dplyr::select(SYMBOL, t) %>% 
  na.omit() %>% 
  group_by(SYMBOL) %>% 
  summarize(t = mean(t)) %>%
  distinct()
res2_high <- limma.res_high %>% 
  dplyr::select(SYMBOL, t) %>% 
  na.omit() %>% 
  group_by(SYMBOL) %>% 
  summarize(t = mean(t)) %>%
  distinct()
```
---GSEA
```{r}
library(fgsea)
library(GSEABase)
library(stats)
packageVersion("fgsea")
ranks_high <- deframe(res2_high)
ranks_low <- deframe(res2_low)
```
--- Getting gene sets from the MsigDB- https://www.gsea-msigdb.org/gsea/index.jsp 
```{r}
pathways.hallmark <- gmtPathways("c2.cp.wikipathways.v2023.1.Hs.symbols (2).gmt")

pathways.hallmark %>% 
  head() %>% 
  lapply(head)

```
--- Running Fgsea, NOTE computer cant do more than 300 permutations
```{r}
fgseaRes_high_pos <- fgsea(pathways=pathways.hallmark, stats=ranks_high,nperm=300,scoreType='pos')
fgseaRes_high_neg <- fgsea(pathways=pathways.hallmark, stats=ranks_high,nperm=300,scoreType='neg')

fgseaRes_low_pos <- fgsea(pathways=pathways.hallmark, stats=ranks_low,nperm=300,scoreType="pos")
fgseaRes_low_neg<- fgsea(pathways=pathways.hallmark, stats=ranks_low
```
--- Creating upregulated and down regulated pathway plots in each condition
```{r}
fgseaResTidy_high_pos <- fgseaRes_high_pos %>%
  as_tibble() %>%
  arrange(desc(NES)),


fgseaResTidy_high_pos %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  DT::datatable(),


fgseaResTidy_high_neg <- fgseaRes_high_neg %>%
  as_tibble() %>%
  arrange(desc(NES)),


fgseaResTidy_high_neg %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  DT::datatable(),



#####NT5Elo
fgseaResTidy_low_pos <- fgseaRes_low_pos%>%
  as_tibble() %>%
  arrange(desc(NES)),


fgseaResTidy_low_pos %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  DT::datatable(),

fgseaResTidy_low_neg <- fgseaRes_low_neg%>%
  as_tibble() %>%
  arrange(desc(NES)),


fgseaResTidy_low_neg %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  DT::datatable()


```
---Creating plots from Fsgea results
```{r}
library(ggplot2)
library(dplyr)

NT5E_low_GSEA_pos <- fgseaResTidy_low_pos %>%
  filter(pval < 0.05) %>%
  arrange(desc(NES)) %>%
  dplyr::slice(1:10) %>%
  ggplot(aes(reorder(pathway, NES), NES)) +
  geom_col(fill='green') +
  coord_flip() +
  labs(
    x = "Pathway",
    y = "Normalized Enrichment Score",
    title = "Top Upregulated Wikipathways NES from GSEA for NT5Elo Lung Adenocarcinoma Samples"
  ) +
  theme_minimal()
NT5E_high_GSEA_neg

NT5E_low_GSEA_pos
NT5E_low_GSEA_neg <- fgseaResTidy_low_neg %>%
  filter(pval < 0.05) %>%
  arrange(desc(NES)) %>%
  dplyr::slice(1:10) %>%
  ggplot(aes(reorder(pathway, NES), NES)) +
  geom_col(fill='orange') +
  coord_flip() +
  labs(
    x = "Pathway",
    y = "Normalized Enrichment Score",
    title = "Top Downregulated Wikipathways NES from GSEA for NT5Elo Lung Adenocarcinoma Samples"
  ) +
  theme_minimal()

NT5E_low_GSEA_neg

print(NT5E_low_GSEA)
NT5E_high_GSEA_pos <- fgseaResTidy_high_pos %>%
  filter(pval < 0.05) %>%
  arrange(desc(NES)) %>%
  dplyr::slice(1:10) %>%
  ggplot(aes(reorder(pathway, NES), NES)) +
  geom_col(fill='green') +
  coord_flip() +
  labs(
    x = "Pathway",
    y = "Normalized Enrichment Score",
    title = "Top Upregulated Wikipathways NES from GSEA for NT5Ehi Lung Adenocarcinoma Samples"
  ) +
  theme_minimal()
NT5E_high_GSEA_neg <- fgseaResTidy_high_neg %>%
  filter(pval < 0.05) %>%
  arrange(desc(NES)) %>%
  dplyr::slice(1:10) %>%
  ggplot(aes(reorder(pathway, NES), NES)) +
  geom_col(fill='orange') +
  coord_flip() +
  labs(
    x = "Pathway",
    y = "Normalized Enrichment Score",
    title = "Top Downregulated Wikipathways NES from GSEA for NT5Ehi Lung Adenocarcinoma Samples"
  ) +
  theme_minimal()
```
---Saving Plots
```{r}
combined_GSEA_plot_NT5Elo <- plot_grid(NT5E_low_GSEA_neg,NT5E_low_GSEA_pos)

combined_GSEA_plot_NT5Ehi <- plot_grid(NT5E_high_GSEA_neg,NT5E_high_GSEA_pos)


save_plot("Wikipathways NT5Ehi pos 30.tiff",NT5E_high_GSEA_pos, base_height = 15,base_width = 15)
save_plot("Wikipathways NT5Ehi neg 30.tiff",NT5E_high_GSEA_neg, base_height = 15,base_width = 15)


save_plot("Wikipathways NT5Elo pos 30.tiff",NT5E_low_GSEA_pos, base_height = 15,base_width = 15)
save_plot("Wikipathways NT5Elo neg 30.tiff",NT5E_low_GSEA_neg, base_height = 15,base_width = 15)

```
---Immune deconvolution
```{r}
library(immunedeconv)
library(EPIC)
library(biomaRt)
library(RColorBrewer)
library(dplyr)
library(ggplot2)
library(tidyr)
library(immunedeconv)
library(tibble)
library(dplyr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(gridExtra)

```
--- Seperating Tumour Samples, renaming emsembl IDs with Gene Symbols
```{r}
Gene_expression_matrix <- limma_res$voomObj[limma_res$voomObj$targets$definition == 'Primary solid Tumor',]
rownames(Gene_expression_matrix$E)<- NULL
rownames(Gene_expression_matrix$E)<- Gene_expression_matrix$genes[,7]
View(Gene_expression_matrix$E)
rownames(Gene_expression_matrix$E)
dim(Gene_expression_matrix$E)
```
--- Defining NT5Ehi and NT5Elo samples
```{r}
###NT5Ehi

upperquartile<-apply(Gene_expression_matrix$E, 1, quantile, probs = 0.773)
samples_above_quartile <- Gene_expression_matrix$E['NT5E',Gene_expression_matrix$targets$definition=='Primary solid Tumor'] > upperquartile['NT5E']
expr_samples_above_quartile_mat <- Gene_expression_matrix$E[,samples_above_quartile]
dim(expr_samples_above_quartile_mat)




##### low expression of NT5E
lowerquartile<-apply(Gene_expression_matrix$E, 1, quantile, probs = 0.25)
samples_below_quartile <- Gene_expression_matrix$E['NT5E',Gene_expression_matrix$targets$definition=='Primary solid Tumor'] < lowerquartile['NT5E']
expr_samples_below_quartile_mat <- Gene_expression_matrix$E[,samples_below_quartile]

```
---Running xCell deconvolution from immundeconv package
```{r}
NT5E_hi_res_xcell <- deconvolute(expr_samples_above_quartile_mat, "xcell", tumor = TRUE)


NT5ELOres_xcell <- deconvolute(expr_samples_below_quartile_mat, "xcell", tumor = TRUE)

```
--- Simplifying results by grouping cell types based on common phenotype characteristics
```{r}
# Combine the data using bind_rows
names(NT5E_hi_res_xcell) <- c("cell_type", as.character(seq_along(NT5E_hi_res_xcell)[-1]))
names(NT5ELOres_xcell) <- c("cell_type", as.character(seq_along(NT5ELOres_xcell)[-1]))

combined_data_long <- rbind(
  NT5E_hi_res_xcell %>% mutate(matrix = "NT5E hi"),
  NT5ELOres_xcell %>% mutate(matrix = "NT5E lo"))
View(combined_data_long)
# Define the cell type groups based on combining rows
cell_type_groups <- c(
  "T Cell CD4+" = c("T cell CD4+ memory", "T cell CD4+ naive", "T cell CD4+ (non-regulatory)", "T cell CD4+ central memory", "T cell CD4+ effector memory", "T cell CD4+ Th1", "T cell CD4+ Th2"),
  "T Cell CD8+" = c("T cell CD8+ naive", "T cell CD8+", "T cell CD8+ central memory", "T cell CD8+ effector memory"),
  "Tregs" = c("T cell regulatory (Tregs)"),
  "Macrophages" = c("Macrophage", "Macrophage M1", "Macrophage M2"),
  "B Cell" = c("Class-switched memory B cell", "B cell", "B cell plasma"),
  "Immune Score" = c("immune score"),
  "NK Cells" = c("T cell NK", "NK cell"),
  "Neutrophil" = c("Neutrophil"),
  "Stromal Score" = c("stroma score"))

combined_data_long$grouped_cell_type <- factor(combined_data_long$cell_type,
                                               levels = unlist(cell_type_groups),
                                               labels = names(cell_type_groups))
#label ID for each row ID 1=NT5Ehi and ID2 =NT5Elo
combined_data_long <- combined_data_long %>%
  mutate(ID = ifelse(row_number() <= 39, "1", "2"))
#Filter the data for the subgroups within NT5Elo and NT5Ehi
NT5Elo_subgroups <- combined_data_long %>%
  filter(ID == "2" & grouped_cell_type %in% names(cell_type_groups))

NT5Ehi_subgroups <- combined_data_long %>%
  filter(ID == "1" & grouped_cell_type %in% names(cell_type_groups))
View(combined_data_long)
# Combine combined subgroups
subgroup_data <- bind_rows(NT5Elo_subgroups, NT5Ehi_subgroups)

# Remove the numeric suffix from grouped_cell_type
subgroup_data$grouped_cell_type <- gsub("\\d+$", "", subgroup_data$grouped_cell_type)



```
---Plotting Immune deonconvolution in beeswarm() and getting comparisons for P.Value
```{r}
comparisons<-list(c("NT5Ehi_B Cell","NT5Elo_B Cell"),
                  c("NT5Ehi_T Cell CD4+" ,"NT5Elo_T Cell CD4+"),
                  c("NT5Ehi_T Cell CD8+","NT5Elo_T Cell CD8+"),
                  c("NT5Ehi_Neutrophil" ,"NT5Elo_Neutrophil" ),
                  c("NT5Ehi_NK Cells","NT5Elo_NK Cells"),
                  c("NT5Ehi_Tregs" ,"NT5Elo_Tregs" ),
                  c("NT5Ehi_Immune Score","NT5Elo_Immune Score"),
                  c("NT5Ehi_Stromal Score","NT5Elo_Stromal Score"),
                  c("NT5Ehi_Macrophages","NT5Elo_Macrophages")
P1<-ggplot(subgroup_data_long, aes(x = grouped_cell_type, y = value,fill=ID,col=ID)) +
geom_quasirandom(dodge.width = 100) +
 scale_x_discrete("Cell Type")+
  ylab("Proportion of Cells") +
  ggtitle("Distribution of Gene Expression") +
  facet_wrap(~ grouped_cell_type, scales = "free")+
  theme(axis.text.x = element_blank()) +
  theme_cowplot() + scale_fill_manual(values = c("NT5Elo" = "blue", "NT5Ehi" = "red")) +
  scale_color_manual(values = c("NT5Elo" = "blue", "NT5Ehi" = "red"))
print(P1)
 # NB: stat_compare_means(data = data_all, comparisons = comparisons, method = "wilcox", paired = FALSE) could only be ran without Facet, so P.values were taken manualy and a table was constructed next to the figure containing all P.values of comparisons.
 
save_plot('BeeswarmDeconvplot.tiff',P1,base_height=20, base_width = 10)

```

